---
title: "Transcriptomics Project"
subtitle: ""
author: "Roa Hanoun & Jero Eyras"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: false
    toc_float: false
    number_sections: false
    theme: cosmo
    highlight: tango
fontsize: 12pt
geometry: margin=1in
---

# Exploratory Data Analysis

## Loading data into R

```{r}
data6 <- read.table("./data/GSE268987_LF_6vsC_6_deg.txt", sep = "\t", quote = "", header = T)
head(data6)

dim(data6)
str(data6)
```

```{r}
data36 <- read.table("./data/GSE268987_LF_48vsC_48_deg.txt", sep = "\t", quote = "", header = T)
head(data36)
```


```{r}
Control_6 <- data6[28:31]
names(Control_6) <- c("C6_r1", "C6_r2", "C6_r3", "C6_r4")
Control_6_Normalized <- data6[36:39]
names(Control_6_Normalized) <- c("C6_fkpm_r1", "C6_fkpm_r2", "C6_fkpm_r3", "C6_fkpm_r4")

Exposed_6 <- data6[24:27]
names(Exposed_6) <- c("E6_r1", "E6_r2", "E6_r3", "E6_r4")
Exposed_6_Normalized <- data6[32:35]
names(Exposed_6_Normalized) <- c("E6_fkpm_r1", "E6_fkpm_r2", "E6_fkpm_r3", "E6_fkpm_r4")

Control_36 <- data36[28:31]
names(Control_36) <- c("C36_r1", "C36_r2", "C36_r3", "C36_r4")
Control_36_Normalized <- data36[36:39]
names(Control_36_Normalized) <- c("C36_fkpm_r1", "C36_fkpm_r2", "C36_fkpm_r3", "C36_fkpm_r4")

Exposed_36 <- data36[24:27]
names(Exposed_36) <- c("E36_r1", "E36_r2", "E36_r3", "E36_r4")
Exposed_36_Normalized <- data36[32:35]
names(Exposed_36_Normalized) <- c("E36_fkpm_r1", "E36_fkpm_r2", "E36_fkpm_r3", "E36_fkpm_r4")
```

```{r}
#names(data6)
```

The data has a large range with the maximum and average values being very far apart. This will create a lot of outliers in the plot which will be interesting later on.

## Visualizing using `boxplot` and `density plot`

```{r message=FALSE, warning=FALSE}
# Load required package
library(tidyr)
library(ggplot2)
library(dplyr)


subdata <- data6[24:31] 


summary(subdata)

subdata2 <- data36[24:31] 

summary(subdata2)

```

**box plot after 6 hours**
```{r}
# Convert to long format
long_df <- subdata %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "value"
  )
library(scales)

myColors <- hue_pal()(2)

ggplot(long_df, aes(x = sample, y = value+1)) + # add 1 to solve log (0) values
  geom_boxplot(col=rep(myColors, each=4)) +
  scale_y_continuous(trans = "log2") +
  labs(
    title = "Box Plot of Raw Counts (log₂ scale) for 6 hours experiment",
    x = "Sample",
    y = "Counts (log₂)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
the sample 7 have the highest counts compered with other samples.
**box plot after 36 hours**
```{r}
# Convert to long format
long_df2 <- subdata2 %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "value"
  )

ggplot(long_df2, aes(x = sample, y = value +1)) + 
  geom_boxplot(col=rep(myColors, each=4)) +
  scale_y_continuous(trans = "log2") +
  labs(
    title = "Box Plot of Raw Counts (log₂ scale) for 36 hours experiment",
    x = "Sample",
    y = "Counts (log₂)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

**Bar plot after 6 hours**

```{r}
ggplot(long_df, aes(x = sample, y = value/10^6)) +
  stat_summary(
    fun = sum, 
    geom = "col", 
    fill =rep(myColors, each=4)

  )  +
  scale_y_continuous(labels = scales::comma)+
  
  labs(
    title = "bar Plot of Raw Counts (/10^6) for 6 hours experiment",
    x = "Sample",
    y = "Counts (/10^6)"
  ) +

  theme_minimal()



```
**Bar plot after 36 hours**
```{r}
ggplot(long_df2, aes(x = sample, y = value/10^6)) +
  stat_summary(
    fun = sum, 
    geom = "col", 
    fill =rep(myColors, each=4)
  )  +
  scale_y_continuous(labels = scales::comma)+
   
  labs(
    title = "bar Plot of Raw Counts (/10^6) for 36 hours experiment",
    x = "Sample",
    y = "Counts (/10^6)"
  ) +

  theme_minimal()

```

**Density plot after 6 hours**

```{r message=FALSE, warning=FALSE}
## The affy library has a density plotting function
library(affy)

## Create a list of 4 colors to use which are the same used throughout this chapter 
library(scales)
myColors <- scales::hue_pal()(2)

## Plot the log2-transformed data with a 0.1 pseudocount
plotDensity(log2(subdata + 0.1), col=rep(myColors, each=4),
            lty=c(1:ncol(subdata)), xlab='Log2(count)',
            main='Expression Distribution')

## Add a legend and vertical line
legend('topright', names(subdata), lty=c(1:ncol(subdata)),
       col=rep(myColors, each=4))
abline(v=-1.5, lwd=1, col='red', lty=2)
```
**Density plot after 36 hours**
```{r}

myColors <- hue_pal()(2)

## Plot the log2-transformed data with a 0.1 pseudocount
plotDensity(log2(subdata2 + 0.1), col=rep(myColors, each=4),
            lty=c(1:ncol(subdata2)), xlab='Log2(count)',
            main='Expression Distribution')

## Add a legend and vertical line
legend('topright', names(subdata2), lty=c(1:ncol(subdata2)),
       col=rep(myColors, each=4))
abline(v=-1.5, lwd=1, col='red', lty=2)
```
**heatmap**

```{r include=FALSE}
if (!require("pheatmap"))
  BiocManager::install("pheatmap")
```

```{r message=FALSE}
library(pheatmap)
```

```{r}
VisualizeNormalizedDistanceHeatMap <- function(data, label) {
  ddsMat <- DESeqDataSetFromMatrix(countData = data,
                                  colData = data.frame(samples = names(data)),
                                  design = ~ 1)
  
  # Perform normalization
  rld.dds <- vst(ddsMat)
  # 'Extract' normalized values
  rld <- assay(rld.dds)
  
  sampleDist <- dist(t(rld))
  sampleDistMatrix <- as.matrix(sampleDist)

  annotation <- data.frame(Samples = factor(rep(1:2, each = 4), labels = c("CTROL", "FLX")))
  rownames(annotation) <- names(data)

  pheatmap(sampleDistMatrix,
           show_colnames = F,
           annotation_col = annotation,
           clustering_distance_rows = sampleDist,
           clustering_distance_cols = sampleDist,
           main = paste("Euclidean Sample Distances", label))
}
VisualizeNormalizedDistanceHeatMap(data.frame(Control_6, Exposed_6), "6h")
VisualizeNormalizedDistanceHeatMap(data.frame(Control_36, Exposed_36), "36h")
```

```{r include=FALSE}
if (!require("PoiClaClu"))
  BiocManager::install("PoiClaClu")
```

```{r message=FALSE}
library(PoiClaClu)
```

```{r include=FALSE}
if (!require("DESeq2"))
  BiocManager::install("DESeq2")
```

```{r message=FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
```

```{r}
library(DESeq2)
```

```{r include=FALSE}
if (!require("ggplot2"))
  BiocManager::install("ggplot2")
```

```{r message=FALSE}
library(ggplot2)
```

```{r}
VisualizePoissonDistance <- function(data, label) {
  ddsMat <- DESeqDataSetFromMatrix(countData = data,
                                  colData = data.frame(samples = names(data)),
                                  design = ~ 1)
  posid <- PoissonDistance(t(assay(ddsMat)), type = "deseq")
  samplePoisDistMatrix <- as.matrix(posid$dd)
  mdsPoisData <- data.frame(cmdscale(samplePoisDistMatrix))
  names(mdsPoisData) <- c('x_coord', 'y_coord')
  
  groups <- factor(rep(1:2, each = 4), labels = c("CTROL", "FLX"))
  colData <- names(data)
  
  ggplot(mdsPoisData, aes(x = x_coord, y = y_coord, color = groups, label = colData)) + 
  geom_text(size = 4) +
  labs(title = paste("Multi Dimensional Scaling, Sample", label), x = "Poisson Distance", y = "Poisson Distance") +
  theme_bw()
}
VisualizePoissonDistance(data.frame(Control_6, Exposed_6), "6h")
VisualizePoissonDistance(data.frame(Control_36, Exposed_36), "36h")
```

```{r}
data_6 <- data.frame(Control_6, Exposed_6)
t <- data.frame(samples = names(data_6), group = factor(rep(1:2, each = 4), labels = c("control", "exposed")))

dds <- DESeqDataSetFromMatrix(countData = data_6,
                                  colData = t,
                                  design = ~ group)

dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res <- results(dds, name="group_exposed_vs_control")
#res <- results(dds, contrast=c("group","control","exposed"))
res
```

```{r}
t <- DataFrame(res)
#t <- drop_na(t)

ggplot(t, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05))
```

```{r}
summary(res)
```