---
title: "Transcriptomics Project"
subtitle: ""
author: "Roa Hanoun & Jero Eyras"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: false
    toc_float: false
    number_sections: false
    theme: cosmo
    highlight: tango
fontsize: 12pt
geometry: margin=1in
---
```{r message=TRUE, warning=FALSE}
# Core tidyverse packages
library(tidyverse)      # ggplot2, dplyr, tidyr, readr, etc.
library(ggplot2)        # plotting
library(scales)         # comma formatting, hue_pal()

# Color palettes
#library(RColorBrewer)

# Bioconductor core packages
library(DESeq2)
library(SummarizedExperiment)
library(BiocGenerics)
library(S4Vectors)
library(IRanges)

# Data reshaping
#library(reshape2)
library(tidyr)
library(dplyr)

## The affy library has a density plotting function
library(affy)

library(pheatmap)
library(PoiClaClu)
```

# Exploratory Data Analysis

## Loading data into R

```{r}
data6 <- read.table("./data/GSE268987_LF_6vsC_6_deg.txt", sep = "\t", quote = "", header = T)
str(data6)
```

```{r}
data36 <- read.table("./data/GSE268987_LF_48vsC_48_deg.txt", sep = "\t", quote = "", header = T)
str(data36)
```

```{r}

gene_ids6 <- data6[, 1]  
gene_ids36 <- data36[, 1]  

Control_6 <- data6[28:31] 
names(Control_6) <- c("C6_r1", "C6_r2", "C6_r3", "C6_r4")

Control_6_Normalized <- data6[36:39]
names(Control_6_Normalized) <- c("C6_fkpm_r1", "C6_fkpm_r2", "C6_fkpm_r3", "C6_fkpm_r4")

Exposed_6 <- data6[24:27]  
names(Exposed_6) <- c("E6_r1", "E6_r2", "E6_r3", "E6_r4")

Exposed_6_Normalized <- data6[32:35]
names(Exposed_6_Normalized) <- c( "E6_fkpm_r1", "E6_fkpm_r2", "E6_fkpm_r3", "E6_fkpm_r4")

Control_36 <- data36[28:31] 
names(Control_36) <- c( "C36_r1", "C36_r2", "C36_r3", "C36_r4")

Control_36_Normalized <- data36[36:39]
names(Control_36_Normalized) <- c( "C36_fkpm_r1", "C36_fkpm_r2", "C36_fkpm_r3", "C36_fkpm_r4")

Exposed_36 <- data36[24:27] 
names(Exposed_36) <- c( "E36_r1", "E36_r2", "E36_r3", "E36_r4")

Exposed_36_Normalized <- data36[32:35]
names(Exposed_36_Normalized) <- c( "E36_fkpm_r1", "E36_fkpm_r2", "E36_fkpm_r3", "E36_fkpm_r4")

data_6 <- data.frame(Control_6, Exposed_6)
rownames(data_6) <- gene_ids6  


data_36 <- data.frame(Control_36, Exposed_36)
rownames(data_36) <- gene_ids36  

```

The data has a large range with the maximum and average values being very far apart. This will create a lot of outliers in the plot which will be interesting later on.

## Visualizing using `boxplot` and `density plot`

### Statistics

```{r}

subdata <- cbind(Exposed_6, Control_6)
#subdata

summary(subdata)

subdata2 <- cbind(Control_36,Exposed_36)
summary(subdata2)

```

### Boxplots

**box plot after 6 hours**
```{r}
# Convert to long format
long_df <- subdata %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "value"
  )
myColors <- hue_pal()(2)

ggplot(long_df, aes(x = sample, y = value+1)) + # add 1 to solve log (0) values
  geom_boxplot(col=rep(myColors, each=4)) +
  scale_y_continuous(trans = "log2") +
  labs(
    title = "Box Plot of Raw Counts (log₂ scale) for 6 hours experiment",
    x = "Sample",
    y = "Counts (log₂)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
the sample E6_r3 have the highest counts compered with other samples.
**box plot after 36 hours**
```{r}
# Convert to long format
long_df2 <- subdata2 %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "value"
  )

ggplot(long_df2, aes(x = sample, y = value +1)) + 
  geom_boxplot(col=rep(myColors, each=4)) +
  scale_y_continuous(trans = "log2") +
  labs(
    title = "Box Plot of Raw Counts (log₂ scale) for 36 hours experiment",
    x = "Sample",
    y = "Counts (log₂)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
### Barplots

**Bar plot after 6 hours**

```{r}
barplot(colSums(subdata)/10^6)
ggplot(long_df, aes(x = sample, y = value/10^6)) +
  stat_summary(
    fun = sum, 
    geom = "col", 
    fill =rep(myColors, each=4)

  )  +
  scale_y_continuous(labels = scales::comma)+
  
  labs(
    title = "bar Plot of Raw Counts (/10^6) for 6 hours experiment",
    x = "Sample",
    y = "Counts (/10^6)"
  ) +

  theme_minimal()
```

**Bar plot after 36 hours**

```{r}
ggplot(long_df2, aes(x = sample, y = value/10^6)) +
  stat_summary(
    fun = sum, 
    geom = "col", 
    fill =rep(myColors, each=4)
  )  +
  scale_y_continuous(labels = scales::comma)+
   
  labs(
    title = "bar Plot of Raw Counts (/10^6) for 36 hours experiment",
    x = "Sample",
    y = "Counts (/10^6)"
  ) +

  theme_minimal()


```

### Density plots

**Density plot after 6 hours**

```{r}

## Create a list of 4 colors to use which are the same used throughout this chapter 
myColors <- scales::hue_pal()(2)

## Plot the log2-transformed data with a 0.1 pseudocount
plotDensity(log2(subdata + 0.1), col=rep(myColors, each=4),
            lty=c(1:ncol(subdata)), xlab='Log2(count)',
            main='Expression Distribution')

## Add a legend and vertical line
legend('topright', names(subdata), lty=c(1:ncol(subdata)),
       col=rep(myColors, each=4))
abline(v=-1.5, lwd=1, col='red', lty=2)
```
**Density plot after 36 hours**
```{r}

myColors <- hue_pal()(2)

## Plot the log2-transformed data with a 0.1 pseudocount
plotDensity(log2(subdata2 + 0.1), col=rep(myColors, each=4),
            lty=c(1:ncol(subdata2)), xlab='Log2(count)',
            main='Expression Distribution')

## Add a legend and vertical line
legend('topright', names(subdata2), lty=c(1:ncol(subdata2)),
       col=rep(myColors, each=4))
abline(v=-1.5, lwd=1, col='red', lty=2)
```
After analyzing the above plots, two samples appear irregular: E6_r3 from the 6-hour dataset (though its impact is minimal) and C36_r1 from the 36-hour dataset. These will be noted for potential removal in the final step of analysis.

## Visualizing using `heatmap` and `MDS`
This section adds a few Exploratory Data Analysis techniques where we will measure and look at distances between samples based on normalized data.
**heatmap**

```{r include=FALSE}
if (!require("pheatmap"))
  BiocManager::install("pheatmap")
```

```{r message=FALSE, warning=FALSE}
#library(pheatmap)
#library(DESeq2)
```
### Normalization

The updated values which are now comparable across genes whereas the raw count data was harder to compare.

```{r}
VisualizeNormalizedDistanceHeatMap <- function(data, label) {
  ddsMat <- DESeqDataSetFromMatrix(countData = data,
                                  colData = data.frame(samples = names(data)),
                                  design = ~ 1)
  
  # Perform normalization
  rld.dds <- vst(ddsMat)
  # 'Extract' normalized values
  rld <- assay(rld.dds)
  
  sampleDist <- dist(t(rld))
  sampleDistMatrix <- as.matrix(sampleDist)

  annotation <- data.frame(Samples = factor(rep(1:2, each = 4), labels = c("CTROL", "FLX")))
  rownames(annotation) <- names(data)

  pheatmap(sampleDistMatrix,
           show_colnames = F,
           annotation_col = annotation,
           clustering_distance_rows = sampleDist,
           clustering_distance_cols = sampleDist,
           main = paste("Euclidean Sample Distances", label))
}
VisualizeNormalizedDistanceHeatMap(data_6, "6h")
VisualizeNormalizedDistanceHeatMap(data_36, "36h")
```

```{r include=FALSE}
if (!require("PoiClaClu"))
  BiocManager::install("PoiClaClu")
```

```{r message=FALSE}
#library(PoiClaClu)
```

```{r include=FALSE}
if (!require("DESeq2"))
  BiocManager::install("DESeq2")
```

```{r message=FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
```

```{r}
#library(DESeq2)
```

```{r include=FALSE}
if (!require("ggplot2"))
  BiocManager::install("ggplot2")
```

```{r message=FALSE}
#library(ggplot2)
```

```{r}
VisualizePoissonDistance <- function(data, label) {
  ddsMat <- DESeqDataSetFromMatrix(countData = data,
                                  colData = data.frame(samples = names(data)),
                                  design = ~ 1)
  posid <- PoissonDistance(t(assay(ddsMat)), type = "deseq")
  samplePoisDistMatrix <- as.matrix(posid$dd)
  mdsPoisData <- data.frame(cmdscale(samplePoisDistMatrix))
  names(mdsPoisData) <- c('x_coord', 'y_coord')
  
  groups <- factor(rep(1:2, each = 4), labels = c("CTROL", "FLX"))
  colData <- names(data)
  
  ggplot(mdsPoisData, aes(x = x_coord, y = y_coord, color = groups, label = colData)) + 
  geom_text(size = 4) +
  labs(title = paste("Multi Dimensional Scaling, Sample", label), x = "Poisson Distance", y = "Poisson Distance") +
  theme_bw()
}
VisualizePoissonDistance(data_6, "6h")
VisualizePoissonDistance(data_36, "36h")
```

After generating heatmaps and MDS plots, the most irregular samples identified are: C36-R1 and E36_R3 from the 36-hour dataset, and C6_R4 from the 6-hour dataset.

### clean data
The only clear rule is that each group must retain at least three samples. If one of the three samples visibly deviates from the other replicates, even after normalization, it will stay in your data set and it should be mentioned in your analysis report/ final article that noise may be introduced by this sample. 
```{r}
Control_6 <- Control_6[-4]  # Removes C6_r4
Control_36 <- Control_36[-1]  # Removes C36_r1
Exposed_36 <- Exposed_36[-3]  # Removes E36_r3

```
# Discovering Differentialy Expressed Genes (DEGs)  

```{r}
data_6 <- data.frame(Control_6, Exposed_6)
rownames(data_6) <- gene_ids6  


t <- data.frame(samples = names(data_6), group = factor(c(rep("control", 3), rep("exposed", 4)), labels = c("control", "exposed")))

dds <- DESeqDataSetFromMatrix(countData = data_6,
                                  colData = t,
                                  design = ~ group)

dds <- DESeq(dds)
```

In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points will be colored blue if the adjusted p value is less than 0.1. Points which fall out of the window are plotted as open triangles pointing either up or down.


```{r}
res <- results(dds)
res <- results(dds, name="group_exposed_vs_control")
res


#plotMA(res)

#resultsNames(dds)

#res <- results(dds, contrast=c("group","exposed","control"))
#res
```

Log fold change shrinkage for visualization and ranking:

“It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.”

```{r message=FALSE, warning=FALSE}
resLFC <- lfcShrink(dds, coef="group_exposed_vs_control", type="apeglm")
resLFC

# For MA plots and visualization
plotMA(resLFC)  # Better than using res

# For volcano plots (use shrunken LFC with original p-values)
plot(resLFC$log2FoldChange, -log10(resLFC$pvalue))

```

```{r}
t <- DataFrame(res)
#t <- drop_na(t)

ggplot(t, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05))

```

```{r}
library(EnhancedVolcano)
EnhancedVolcano(res, x = 'log2FoldChange', y = 'pvalue',
                         lab=rownames(res),
                         #title = paste(data_6, "control vs exposed"),
                         subtitle = bquote(italic('FDR <= 0.05 and absolute FC >= 2')),
                         # Change text and icon sizes
                       labSize = 3, pointSize = 1.5, axisLabSize=10, titleLabSize=12,
                         subtitleLabSize=8, captionLabSize=10,
                         # Disable legend
                         legendPosition = "none",
                         # Set cutoffs
                          ylim = c(0,4),
                         pCutoff = 0.05, FCcutoff = 0)


## Note that the most simplest case is:
EnhancedVolcano(res, x = 'log2FoldChange', y = 'pvalue', lab = rownames(res))


```

```{r}
#Here we specify the gene which had the smallest p value from the results table created above. You can select the gene to plot by rowname or by numeric index.
plotCounts(dds, gene=which.min(res$padj), intgroup="group")

#mcols(res)$description



```

```{r}
#summary(res)
s <- data.frame(t)
s <- filter(s , pvalue < 0.05)



length(s$pvalue)
89 + 76
```

```{r}
ggplot(t, aes(x = round(pvalue, digits = 2))) +
  geom_bar() +
  geom_vline(xintercept = 0.05, colour = "red")
```