---
title: "Transcriptomics Project"
subtitle: ""
author: "Roa Hanoun & Jero Eyras"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: false
    toc_float: false
    number_sections: false
    theme: cosmo
    highlight: tango
fontsize: 12pt
geometry: margin=1in
---

# Exploratory Data Analysis

## Loading data into R

```{r}
data6 <- read.table("./data/GSE268987_LF_6vsC_6_deg.txt", sep = "\t", quote = "", header = T)
head(data6)

dim(data6)
str(data6)
```

```{r}
data36 <- read.table("./data/GSE268987_LF_48vsC_48_deg.txt", sep = "\t", quote = "", header = T)
head(data36)
```


```{r}
Control_6 <- data6[28:31] 
Control_6_Normalized <- data6[36:39]

Exposed_6 <- data6[24:27] 
Exposed_6_Normalized <- data6[32:35]

Control_36 <- data36[28:31] 
Control_36_Normalized <- data36[36:39]

Exposed_36 <- data36[24:27]
Exposed_36_Normalized <- data36[32:35]
```

```{r}
#names(data6)
```

The data has a large range with the maximum and average values being very far apart. This will create a lot of outliers in the plot which will be interesting later on.

## Visualizing using `boxplot` and `density plot`

```{r}
# Load required package
library(tidyr)
library(ggplot2)
library(dplyr)

samples <- c("Q05_count" , "Q06_count" ,"Q07_count","Q08_count",
             "Q01_count","Q02_count","Q03_count","Q04_count")

(subdata <- data6[samples])

summary(subdata)

samples2 <- c("Q13_count" , "Q14_count" ,"Q15_count","Q16_count",
             "Q09_count","Q10_count","Q11_count","Q12_count")

subdata2 <- data36[samples2]
summary(subdata2)

```

**box plot after 6 hours**
```{r}
# Convert to long format
long_df <- subdata %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "value"
  )

ggplot(long_df, aes(x = sample, y = value+1, fill = sample)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log2") +
 # ylim(0, 600) + 
  labs(
    title = "Box Plot of Raw Counts (log₂ scale)",
    x = "Sample",
    y = "Counts (log₂)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
**box plot after 36 hours**
```{r}
# Convert to long format
long_df2 <- subdata2 %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "value"
  )

ggplot(long_df2, aes(x = sample, y = value +1, fill = sample)) + # i added 1 to the y values to keep the 0 values ! 
  geom_boxplot() +
  scale_y_continuous(trans = "log2") +
 # ylim(0, 600) + 
  labs(
    title = "Box Plot of Raw Counts (log₂ scale)",
    x = "Sample",
    y = "Counts (log₂)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

**Bar plot after 6 hours**
```{r}
ggplot(long_df, aes(x = sample, y = value)) +
  stat_summary(
    fun = sum, 
    geom = "col", 
    fill = "tomato"
  )  +
  scale_y_continuous(labels = scales::comma)+

  theme_minimal()

```
**Bar plot after 36 hours**
```{r}
ggplot(long_df2, aes(x = sample, y = value)) +
  stat_summary(
    fun = sum, 
    geom = "col", 
    fill = "tomato"
  )  +
  scale_y_continuous(labels = scales::comma)+

  theme_minimal()

```

**Density plot after 6 hours**

```{r}
## The affy library has a density plotting function
library(affy)

## Create a list of 4 colors to use which are the same used throughout this chapter 
library(scales)
myColors <- hue_pal()(2)

## Plot the log2-transformed data with a 0.1 pseudocount
plotDensity(log2(subdata + 0.1), col=rep(myColors, each=4),
            lty=c(1:ncol(subdata)), xlab='Log2(count)',
            main='Expression Distribution')

## Add a legend and vertical line
legend('topright', names(subdata), lty=c(1:ncol(subdata)),
       col=rep(myColors, each=4))
abline(v=-1.5, lwd=1, col='red', lty=2)
```
**Density plot after 36 hours**
```{r}

myColors <- hue_pal()(2)

## Plot the log2-transformed data with a 0.1 pseudocount
plotDensity(log2(subdata2 + 0.1), col=rep(myColors, each=4),
            lty=c(1:ncol(subdata2)), xlab='Log2(count)',
            main='Expression Distribution')

## Add a legend and vertical line
legend('topright', names(subdata2), lty=c(1:ncol(subdata2)),
       col=rep(myColors, each=4))
abline(v=-1.5, lwd=1, col='red', lty=2)
```
**heatmap**

```{r include=FALSE}
if (!require("pheatmap"))
  BiocManager::install("pheatmap")
```

```{r message=FALSE}
library(pheatmap)
```

```{r}
VisualizeNormalizedDistanceHeatMap <- function(data, label) {
  sampleDist <- dist(t(data))
  sampleDistMatrix <- as.matrix(sampleDist)

  annotation <- data.frame(Samples = factor(rep(1:2, each = 4), labels = c("FLX", "CTROL")))
  rownames(annotation) <- names(data)

  pheatmap(sampleDistMatrix,
           show_colnames = F,
           annotation_col = annotation,
           clustering_distance_rows = sampleDist,
           clustering_distance_cols = sampleDist,
           main = paste("Euclidean Sample Distances", label))
}
VisualizeNormalizedDistanceHeatMap(data6[32:39], "6h")
VisualizeNormalizedDistanceHeatMap(data36[32:39], "36h")
```

```{r include=FALSE}
if (!require("PoiClaClu"))
  BiocManager::install("PoiClaClu")
```

```{r message=FALSE}
library(PoiClaClu)
```

```{r include=FALSE}
if (!require("DESeq2"))
  BiocManager::install("DESeq2")
```

```{r message=FALSE}
library(DESeq2)
```

```{r include=FALSE}
if (!require("ggplot2"))
  BiocManager::install("ggplot2")
```

```{r message=FALSE}
library(ggplot2)
```

```{r}
VisualizePoissonDistance <- function(data, label) {
  ddsMat <- DESeqDataSetFromMatrix(countData = data,
                                  colData = data.frame(samples = names(data)),
                                  design = ~ 1)
  posid <- PoissonDistance(t(assay(ddsMat)), type = "deseq")
  samplePoisDistMatrix <- as.matrix(posid$dd)
  mdsPoisData <- data.frame(cmdscale(samplePoisDistMatrix))
  names(mdsPoisData) <- c('x_coord', 'y_coord')
  
  groups <- factor(rep(1:2, each = 4), labels = c("FLX", "CTROL"))
  colData <- names(data)
  
  ggplot(mdsPoisData, aes(x = x_coord, y = y_coord, color = groups, label = colData)) + 
  geom_text(size = 4) +
  labs(title = paste("Multi Dimensional Scaling, Sample", label), x = "Poisson Distance", y = "Poisson Distance") +
  theme_bw()
}
VisualizePoissonDistance(data6[24:31], "6h")
VisualizePoissonDistance(data36[24:31], "36h")
```
