---
title: "Transcriptomics Project"
subtitle: ""
author: "Roa Hanoun & Jero Eyras"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: false
    toc_float: false
    number_sections: false
    theme: cosmo
    highlight: tango
fontsize: 12pt
geometry: margin=1in
---

```{r message=FALSE}
# Core tidyverse packages
library(tidyverse)      # ggplot2, dplyr, tidyr, readr, etc.
library(ggplot2)        # plotting
library(scales)         # comma formatting, hue_pal()

# Color palettes
#library(RColorBrewer)

# Bioconductor core packages
library(DESeq2)
library(SummarizedExperiment)
library(BiocGenerics)
library(S4Vectors)
library(IRanges)

# Data reshaping
#library(reshape2)
library(tidyr)
library(dplyr)

## The affy library has a density plotting function
library(affy)

library(pheatmap)
library(PoiClaClu)
```

# Exploratory Data Analysis

## Loading data into R

```{r}
data6 <- read.table("./data/GSE268987_LF_6vsC_6_deg.txt", sep = "\t", quote = "", header = T)

control_6 <- data6[28:31] 
names(control_6) <- c("C6_r1", "C6_r2", "C6_r3", "C6_r4")

exposed_6 <- data6[24:27]  
names(exposed_6) <- c("E6_r1", "E6_r2", "E6_r3", "E6_r4")

data_6 <- data.frame(control_6, exposed_6)
rownames(data_6) <- data6[, 1]

head(data_6)
```

```{r}
data36 <- read.table("./data/GSE268987_LF_48vsC_48_deg.txt", sep = "\t", quote = "", header = T)

control_36 <- data36[28:31] 
names(control_36) <- c( "C36_r1", "C36_r2", "C36_r3", "C36_r4")

exposed_36 <- data36[24:27] 
names(exposed_36) <- c( "E36_r1", "E36_r2", "E36_r3", "E36_r4")

data_36 <- data.frame(control_36, exposed_36)
rownames(data_36) <- data36[, 1]

head(data_36)
```

The data has a large range with the maximum and average values being very far apart. This will create a lot of outliers in the plot which will be interesting later on.

## Visualizing using `boxplot` and `density plot`

### Statistics

```{r}
summary(data_6)
```
```{r}
summary(data_36)
```

### Boxplots

```{r}
BoxPlotRawCount <- function(data, label) {
  # Convert to long format
  long_df <- pivot_longer(data,
      cols = everything(),
      names_to = "sample",
      values_to = "value"
    )
  
  myColors <- hue_pal()(2)
  
  ggplot(long_df, aes(x = sample, y = value+1)) + # add 1 to solve log (0) values
  geom_boxplot(col=rep(myColors, each=4)) +
  scale_y_continuous(trans = "log2") +
  labs(
    title = paste("Box Plot of Raw Counts (log₂ scale) for", label, "hours experiment"),
    x = "Sample",
    y = "Counts (log₂)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```
```{r}
#Boxplot 6 hours
BoxPlotRawCount(data_6, "6")
#Boxplot 36 hours
BoxPlotRawCount(data_36, "36")
```

the sample E6_r3 have the highest counts compered with other samples.

### Barplots

```{r}
BarPlotRawCount <- function(data, label) {
  # Convert to long format
  long_df <- pivot_longer(data,
      cols = everything(),
      names_to = "sample",
      values_to = "value"
    )
  
  myColors <- hue_pal()(2)
  
  ggplot(long_df, aes(x = sample, y = value/10^6)) +
    stat_summary(
      fun = sum, 
      geom = "col", 
      fill =rep(myColors, each=4)
    )  +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = paste("Bar Plot of Raw Counts (/10^6) for ", label, "hours experiment"),
      x = "Sample",
      y = "Counts (/10^6)"
    ) +
    theme_minimal()
}
```
```{r}
#Barplot 6 hours
BarPlotRawCount(data_6, "6")
#Barplot 36 hours
BarPlotRawCount(data_36, "36")
```

### Density plots

```{r}
DensityPlotRawCount <- function(data) {
  ## Create a list of 4 colors to use which are the same used throughout this chapter 
  myColors <- hue_pal()(2)
  
  ## Plot the log2-transformed data with a 0.1 pseudocount
  plotDensity(log2(data + 0.1), col=rep(myColors, each=4),
              lty=c(1:ncol(data)), xlab='Log2(count)',
              main='Expression Distribution')
  
  ## Add a legend and vertical line
  legend('topright', names(data), lty=c(1:ncol(data)),
         col=rep(myColors, each=4))
  abline(v=-1.5, lwd=1, col='red', lty=2)
}
```
```{r}
#Density Plot 6 hours
DensityPlotRawCount(data_6)
#Density Plot 36 hours
DensityPlotRawCount(data_36)
```

After analyzing the above plots, two samples appear irregular: E6_r3 from the 6-hour dataset (though its impact is minimal) and C36_r1 from the 36-hour dataset. These will be noted for potential removal in the final step of analysis.

## Visualizing using `heatmap` and `MDS`
This section adds a few Exploratory Data Analysis techniques where we will measure and look at distances between samples based on normalized data.
**heatmap**

### Normalized Distance Heatmap

The updated values which are now comparable across genes whereas the raw count data was harder to compare.

```{r}
VisualizeNormalizedDistanceHeatMap <- function(data, label) {
  ddsMat <- DESeqDataSetFromMatrix(countData = data,
                                  colData = data.frame(samples = names(data)),
                                  design = ~ 1)
  
  # Perform normalization
  rld.dds <- vst(ddsMat)
  # 'Extract' normalized values
  rld <- assay(rld.dds)
  
  sampleDist <- dist(t(rld))
  sampleDistMatrix <- as.matrix(sampleDist)

  annotation <- data.frame(Samples = factor(rep(1:2, each = 4), labels = c("CTROL", "FLX")))
  rownames(annotation) <- names(data)

  pheatmap(sampleDistMatrix,
           show_colnames = F,
           annotation_col = annotation,
           clustering_distance_rows = sampleDist,
           clustering_distance_cols = sampleDist,
           main = paste("Euclidean Sample Distances", label))
}
```
```{r}
#Normalized distance heat map 6 hours
VisualizeNormalizedDistanceHeatMap(data_6, "6 hours")
#Normalized distance heat map 36 hours
VisualizeNormalizedDistanceHeatMap(data_36, "36 hours")
```

### Poisson Distance

```{r}
VisualizePoissonDistance <- function(data, label) {
  ddsMat <- DESeqDataSetFromMatrix(countData = data,
                                  colData = data.frame(samples = names(data)),
                                  design = ~ 1)
  
  #Poisson function already performs normalization
  posid <- PoissonDistance(t(assay(ddsMat)), type = "deseq")
  samplePoisDistMatrix <- as.matrix(posid$dd)
  mdsPoisData <- data.frame(cmdscale(samplePoisDistMatrix))
  names(mdsPoisData) <- c('x_coord', 'y_coord')
  
  groups <- factor(rep(1:2, each = 4), labels = c("CTROL", "FLX"))
  colData <- names(data)
  
  ggplot(mdsPoisData, aes(x = x_coord, y = y_coord, color = groups, label = colData)) + 
  geom_text(size = 4) +
  labs(title = paste("Multi Dimensional Scaling, Sample", label), x = "Poisson Distance", y = "Poisson Distance") +
  theme_bw()
}
```
```{r}
#Poisson Distance 6 hours
VisualizePoissonDistance(data_6, "6 hours")
#Poisson Distance 36 hours
VisualizePoissonDistance(data_36, "36 hours")
```

After generating heatmaps and MDS plots, the most irregular samples identified are: C36-R1 and E36_R3 from the 36-hour dataset, and C6_R4 from the 6-hour dataset.

## Clean data

The only clear rule is that each group must retain at least three samples. If one of the three samples visibly deviates from the other replicates, even after normalization, it will stay in your data set and it should be mentioned in your analysis report/ final article that noise may be introduced by this sample.

```{r}
data_6_clean <- select(data_6, -c("C6_r4")) # Removes C6_r4
data_36_clean <- select(data_36, -c("C36_r1", "E36_r3")) # Removes C36_r1 & E36_r3
```

# Discovering Differentialy Expressed Genes (DEGs)

```{r}
DiscoverDEGsWithDESeq <- function(data, controlSamples, exposedSamples) {
  col <- data.frame(samples = names(data), group = factor(c(rep("control", controlSamples), rep("exposed", exposedSamples)), labels = c("control", "exposed")))
  
  dds <- DESeqDataSetFromMatrix(countData = data,
                                    colData = col,
                                    design = ~ group)
  dds <- DESeq(dds)
  
  res <- results(dds)
  res <- results(dds, name="group_exposed_vs_control")
  resLFC <- lfcShrink(dds, coef="group_exposed_vs_control", type="apeglm")
  
  return(list(deg = res, deg_lfcShrink = resLFC))
}
```
```{r}
#Differential expressed genes 6 hours
deg_6 <- DiscoverDEGsWithDESeq(data_6_clean, 3, 4)
summary(deg_6$deg)
#Differential expressed genes 36 hours
deg_36 <- DiscoverDEGsWithDESeq(data_36_clean, 3, 3)
summary(deg_36$deg)
```

In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points will be colored blue if the adjusted p value is less than 0.1. Points which fall out of the window are plotted as open triangles pointing either up or down.

Log fold change shrinkage for visualization and ranking:

“It is more useful visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.”

```{r message=FALSE, warning=FALSE}
PlotMA <- function(deg) {
  # For MA plots and visualization
  plotMA(deg)  # Better than using res
  
  # For volcano plots (use shrunken LFC with original p-values)
  plot(deg$log2FoldChange, -log10(deg$pvalue))
}
```
```{r}
#Plot LFC 6 hours
PlotMA(deg_6$deg_lfcShrink)
#Plot LFC 36 hours
PlotMA(deg_36$deg_lfcShrink)
```

```{r}
PlotPVal <- function(deg, label) {
  t <- data.frame(deg)
  t <- drop_na(t)
  
  ggplot(t, aes(x = round(pvalue, digits = 2))) +
    geom_bar() +
    geom_vline(xintercept = 0.05, colour = "red") +
    labs(title = paste("PValue Distribution", label))
}
```
```{r}
#pvalue plot 6 hours
PlotPVal(deg_6$deg, "6 hours")
#pvalue plot 36 hours
PlotPVal(deg_36$deg, "36 hours")
```

```{r}
VolcanoPlot <- function(deg, label) {
  t <- data.frame(deg)
  t <- drop_na(t)

  ggplot(t, aes(x = log2FoldChange, y = -log10(pvalue))) +
    geom_point() +
    geom_hline(yintercept = -log10(0.05)) +
    labs(title = paste("Volcano Plot", label))
}
```
```{r}
#Volcano plot 6 hours
VolcanoPlot(deg_6$deg, "6 hours")
#Volcano plot 36 hours
VolcanoPlot(deg_36$deg, "36 hours")
```

!!! log10 of 0.0000000001 = of 10, is this an outlier?

```{r}
FilterDEGs <- function(deg) {
  s <- data.frame(deg)
  return(filter(s , pvalue < 0.05))
}
```
```{r}
#Filtered DEGs 6 hours
deg_6_filtered <- FilterDEGs(deg_6$deg)
nrow(deg_6_filtered)
#Filtered DEGs 36 hours
deg_36_filtered <- FilterDEGs(deg_36$deg)
nrow(deg_36_filtered)
```

```{r include=FALSE}
#if (!require("EnhancedVolcano"))
  #BiocManager::install("EnhancedVolcano")
```

```{r}
#library(EnhancedVolcano)
#EnhancedVolcano(res, x = 'log2FoldChange', y = 'pvalue',
                         #lab=rownames(res),
                         #title = paste(data_6, "control vs exposed"),
                         #subtitle = bquote(italic('FDR <= 0.05 and absolute FC >= 2')),
                         # Change text and icon sizes
                         #labSize = 3, pointSize = 1.5, axisLabSize=10, titleLabSize=12,
                         #subtitleLabSize=8, captionLabSize=10,
                         # Disable legend
                         #legendPosition = "none",
                         # Set cutoffs
                         #ylim = c(0,4),
                         #pCutoff = 0.05, FCcutoff = 0)


## Note that the most simplest case is:
#EnhancedVolcano(res, x = 'log2FoldChange', y = 'padj', lab = rownames(res))
```

```{r}
#Here we specify the gene which had the smallest p value from the results table created above. You can select the gene to plot by rowname or by numeric index.
#plotCounts(dds, gene=which.min(res$padj), intgroup="group")

#mcols(res)$description

```